---
title: "Lab 17"
output: html_document
date: '2022-03-14'
---

# Lab 17

In this lab, we will practice Object Oriented Programming (OOP) in R.

First, let us have a look at `density` function in R. `density(x)` function returns **an object** representing the probability density function estimated from your dataset `x`.

```{r}
x <- rnorm(1000) # generate random samples from a standeard normal distribution
d <- density(x) # estimate a density funciton from our data
print(d) # print out some basic info of the denisty
```

Let us try to plot the density.

```{r}
plot(d)
```

How does it work? By using polymorphism (see lecture slides), the `plot` function recognizes you are trying to plot a density function and generates a density plot for you.

## A Distribution Class

Now let us implement something similar using OOP. We would like to create a `distribution` class. Objects in this class should **contain three variables**: a dataset of random samples, the mean and variance of the dataset. Below is an example:

```{r}
x <- rnorm(1000)
distr <- list(dataset = x, mean = mean(x), variance = var(x))
class(distr) <- "distribution"
```

Now write a function `distribution`, takes one input `x` containing your random samples and returns an **object** from the aforementioned `distribution` class.

```{r}
x <- rnorm(1000)
distribution <- function(x){
  # write your code here
  distr <- list(dataset = x, mean = mean(x), variance = var(x))
  class(distr) <- "distribution"
  return(distr)
}

distr <- distribution(x)
```

Check the class of `distr` using `class` function.

```{r}
# your code here
class(distr)
```

## Print Distribution

If you type `print(distr)`, you will see a large amount of output occupying the whole screen. This is not nice! Let us write a print function that is more concise.

Write `print` function that is **specific for** `distribution` **class**, so that if I type `print(distr`) I get the following output on the screen:

"My dataset contains X observations, with mean Y and variance Z", where X, Y and Z are values depending on my dataset.

**Hint:** Use `paste` function to concatenate strings and numbers.

```{r}
# Your code here
print.distribution <- function(d){
  print(paste("your dataset contains", length(d$dataset), "observations, with mean", d$mean, "and variance", d$variance))
}

print(distr)
```

## Gaussian Distirbution

Now, let us construct an object for a new class `gaussian_distribution`, which **inherits from** `distribution`. Objects in `gaussian_distribution` contain all variables (dataset, mean and variance) in `distribution` class. They also contain two **additional variables**: `mu` and `sigma` corresponding to parameters in Gaussian distribution.

For example, I can create an object from `gaussian_distribution` class using the following commands:

```{r}
distr <- list(dataset = x, mean = mean(x), variance = var(x))
distr$mu <- distr$mean
distr$sigma <- sqrt(distr$variance)
class(distr) <- c("gaussian_distribution", "distribution")
class(distr)
```

Now, write a function `gaussian_distribution` which takes an input `x` containing random samples in your dataset and returns an object from `gaussian_distribution` class.

```{r}
gaussian_distribution <- function(x){
  #write your code here
  distr <- distribution(x)
  distr$mu <- distr$mean
  distr$sigma <- sqrt(distr$variance)
  class(distr) <- c("gaussian_distribution", "distribution")
  return(distr)
}

distr <- gaussian_distribution(x)
class(distr)
```

Since `gaussian_distribution` inherits from `distribution`, the `print` function we wrote for `distribution` class still works for objects from `gaussian_distribution`.

```{r}
print(distr)
```

Let us use polymorphism to change the behavior of `plot` function so that when I type `plot(distr)`, it will produce a Gaussian density plot estimated from my dataset.

**Hint**: `dnorm(x,mu,sigma)` evaluates the density function of a Gaussian distribution with `mu` and `sigma` at position `x`. See `?dnorm` for more. Check out lecture14 for the usages of plotting functions.

```{r}
plot.gaussian_distribution <- function(d){
  #generate numbers from -5, 5 with interval .1
  x <- seq(-5,5,.1)
  # write your code here  
  y <- dnorm(x, d$mu, d$sigma)
  plot(c(-5,5), c(-0,.5), type = "n")
  lines(x,y)
}

plot(distr)
```

## Poisson Distribution

Now, try it yourself.

Create a function `poisson_distribution(x)` that construct an object from class `poisson_distribution` which **inherits from** `distribution`. Objects of `poisson_distribution` class contain all variables (dataset, mean and variance) in `distribution` class. They also contain **one** **additional variable:** `lambda`, the parameter of a Poisson distribution.

**Hint:** How do you estimate `lambda` from your dataset?

```{r}
# write your code here 
poisson_distribution <-  function(x){
  d <- distribution(x)
  d$lambda <- d$mean
  class(d) <- c("poisson_distribution", "distribution")
  return(d)
}

x <- rpois(1000,lambda =2)
distr <- poisson_distribution(x)
class(distr)
```

Use polymorphism to change the behavior of `plot` function so that when I type `plot(distr)`, it will produce a plot of Poisson **probability mass function** estimated from my dataset.

**Hint:** Poisson probability mass function is a discrete function, not a continuous curve.

```{r}
# write your code here 
plot.poisson_distribution <- function(d){
  plot(c(0,10), c(0, .5), type="n")
  x <- 0:10
  y <- dpois(x, lambda = d$lambda)
  points(x, y)
}

plot(distr)
```
