---
title: "R Notebook"
output: html_notebook
---

# Lab 15

In this lab, we will run experiments using R notebook, which is a powerful programming and note taking tool that allows seamless integration between code and documentation. Here is an example:

```{r}
print("hello world!")
```

You may have seen Python notebook in TB1. R notebook is exactly the same idea, but designed for R programming language.

## R markdown

[Markdown](https://en.wikipedia.org/wiki/Markdown) is a popular scripting language that allows you to write **formatted text** in *plain text editors* (such as [notepad]{.underline}) and render it later. You can toggle between the source code and visualization using the "source" and "visual" in the toolbar and see the differences.

Without further ado, let us begin!

## Targets for today

In today's lab, you will apply **functional programming** to code the pairwise distance function that we have encountered before many times. Specifically, given matrices $A \in \mathbb{R}^{n_1 \times m}$ and $B \in \mathbb{R}^{n_2 \times m}$, we would like to construct a matrix $D \in \mathbb{R}^{n_1 \times n_2}$, where $D_{i,j} = \mathrm{dist}(A_{[i,]}, B_{[j,]})$,$A_{[i,]}$ is $i$-th row of $A$ and $$\mathrm{dist}(a,b) = \sqrt{\sum_i (a_i - b_i)^2}.$$

## Preparation

Before you start, I strongly recommend you review [this week's lecture slides](https://github.com/anewgithubname/MATH10017/blob/main/lecs/lec14.pdf) on list and functional programming.

**Write code below,** generating two random matrices $A$ and $B$ with 3 rows and 2 columns which are both filled with observations from standard normal distribution (Hint: `?rnorm`).

```{r}
#type your code here and run it by clicking the green arrow on the top-right corner.


```

Copy `pdist4` function from Lab 13, and compute the distance matrix $D$ using $A$ and $B$. We will use this result for verification.\

```{r}

# Copy the pdist code here


# type your code here

```

## Pairwise Distance

We already know how to write vectorized code for computing $D$. Now let us inspect at this computation from a functional programming point of view.

-   Given two data matrices $A \in \mathbb{R}^{n_1 \times m}$ and $B\in \mathbb{R}^{n_2 \times m}$,
-   Fix $k$, and compute$(A_{[i,k]} - B_{[j,k]})^2$ for all $i, j$.
-   Store the outcome at the $i,j$-th element of a matrix $D_k$.
-   Apply the above computation for all $k$.
-   $D := \sqrt{\sum_k D_k}$

### Task 1

Write a function `diff2` that takes two scalars $a,b$, and compute $(a-b)^2$:

```{r}
#type your code here
```

`outer(X,Y,FUN)` applies function `FUN` two all pairs of elements in `X` and `Y`. In other words, it computes a matrix $D$, $D_{i,j} = \mathrm{fun}(X_i, Y_j), \forall i,j$.

Now use `outer` and `diff2`, compute a matrix $D^{(1)}$, $D_{i,j}^{(1)} = \mathrm{diff2}(A_{[i,1]}, B_{[j,1]})$.

```{r}
#type your code here
```

### Task 2

We would like to repeat the above operation for all $k$. We can use a for loop, but in functional programming, we tend to think we "apply" a data operation to **the** **entire dataset** rather than create loops that iterate over each piece of our dataset.

This is exactly what we did above: We applied `diff2` to all pairs of $i,j$.

Let us create another function `D_k`, that takes one input $k$ and compute $D_k$ using `outer`.

```{r}
# type your code here

```

Now apply this function to a list of numbers (hint: `?lapply`). The list contains integers from 1 to $m$, where $m$ is 2 in this toy case.

```{r}
# type your code here

```

You should get a list of matrices whose $k$ -th element is $D_k$.

### Task 3

Now, all that is left is summing $D_k$ together and we are done! Can we use a for loop to sum all $D_k$ matrices together?

In Functional Programming, we tend to think our program is a huge pipeline that applies functions repeatedly to our data: `op3(op2(op1(data)))...` (recall the "data pipeline" in the lecture slides)

We can rewrite the summation of $D_k$ using this paradigm:

$\sum_{k=1…4} D_k$ is the same as

`sum2(sum2(sum2(D_1,D_2),D_3),D_4)`, where `sum2(X,Y)` sums two matrices `X` and `Y` together. Write a function `sum2` that sums over two matrices

```{r}
# type your code here

```

Now we need to apply this function to the list of $D_k$ recursively. This can be done using a function called `Reduce`. `Reduce(f,x)` applies `f` to a list `x` recursively. In other words, it computes

`f(f(f(x_1, x_2),x_3),x_4)…`

```{r}
# type your code here

```

Square root the outcome.

```{r}
# type your code here

```

### Task 4

Check, is this the same result you obtained using `pdist4`?

```{r}
# type your code here

```

Now, using functional programming tricks we have learned above, can you simplify the computation to just one line of code. Hint: when passing a function `f` to another function `g` as an input argument, we can write the function definition directly in the function all:

```{r}

f <- function(a){
  print(a)
}

g <- function(f){
  print("g calls f with input 1")
  f(1)
}

g(f)

# the above code is exactly the same as the following

g <- function(f){
  print("g calls f with input 1")
  f(1)
}

g(function(a) print(a))

```

Compare your functional code with the implementation of `pdist4`.

Functional programming can greatly simplify our program indeed!
